---
title: "exploring_gasoline"
output: 
  html_document:
    toc: true
    toc_float: true
date: "2026-04-25"
---

## Exploring dataset


```{r}
serie=window(ts(read.table("Gasolina.dat"),start=1993,freq=12))
```


```{r}
plot(serie, main="Gasoline consumption in Spain", ylab="Thousands of tons")
abline(v=1993:2020,lty=3,col=4)
abline(h= 600)
```


```{r}
mean(serie)
sd(serie)
```

```{r}
boxplot(serie~floor(time(serie)))
```


```{r}
m=apply(matrix(serie,nr=12),2,mean)
v=apply(matrix(serie,nr=12),2,var)
plot(m,v,xlab="Medias anuales",ylab="Varianzas anuales",main="serie")
abline(lm(v~m),col=2,lty=3,lwd=2)
```


```{r}
lnserie <- log(serie)

monthplot(lnserie)
```


```{r}
d12lnserie <- diff(lnserie,12)
plot(d12lnserie)
abline(h= mean(d12lnserie), col= 2)

```


```{r}
plot(decompose(lnserie))
```

```{r}

```



```{r}
var(serie)
var(lnserie)
var(d12lnserie)
var(diff(d12lnserie))
```

```{r}
d1d12lnserie <- diff(d12lnserie)
mean(d12lnserie)
mean(d1d12lnserie)
```

```{r}
plot(d1d12lnserie)
```

```{r}
var(d1d12lnserie)
var(diff(d1d12lnserie)) 
```


### Possible models

```{r}
par(mfrow = c(1,2))
acf(d1d12lnserie, main = "ACF", ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d1d12lnserie, main = "PACF", ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
```






## Models 

Seasonal part: ARMA (1,1)

Regular part: 

- MA(1)
- ARMA(1,1)
- MA(2)
- ARMA(1,2)

*Explain why those 4 models were selected*



```{r}
#################Validation#################################
library(tseries)
validation=function(model,dades){
  s=frequency(get(model$series))
  resid=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resid,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resid),3*sd(resid)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resid)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resid)
  qqline(resid,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resid,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resid),sd=sd(resid)),col=2,add=T)
  
  
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resid,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #ACF & PACF of square residuals 
  par(mfrow=c(1,2))
  acf(resid^2,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid^2,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=36)
  names(psis)=paste("psi",1:36)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:20])
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=36)
  names(pis)=paste("pi",1:36)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:20])
  
  ## Add here complementary tests (use with caution!)
  ##---------------------------------------------------------
  cat("\nNormality Tests\n")
  cat("\n--------------------\n")
 
  ##Shapiro-Wilks Normality test
  print(shapiro.test(resid(model)))

  suppressMessages(require(nortest,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  print(ad.test(resid(model)))
  
  suppressMessages(require(tseries,quietly=TRUE,warn.conflicts=FALSE))
  ##Jarque-Bera test
  print(jarque.bera.test(resid(model)))
  
  cat("\nHomoscedasticity Test\n")
  cat("\n--------------------\n")
  suppressMessages(require(lmtest,quietly=TRUE,warn.conflicts=FALSE))
  ##Breusch-Pagan test
  obs=get(model$series)
  print(bptest(resid(model)~I(obs-resid(model))))
  
  cat("\nIndependence Tests\n")
  cat("\n--------------------\n")
  
  ##Durbin-Watson test
  print(dwtest(resid(model)~I(1:length(resid(model)))))
  
  ##Ljung-Box test
  cat("\nLjung-Box test\n")
  print(t(apply(matrix(c(1:4,(1:4)*s)),1,function(el) {
    te=Box.test(resid(model),type="Ljung-Box",lag=el)
    c(lag=(te$parameter),statistic=te$statistic[[1]],p.value=te$p.value)})))
  

  #Sample ACF vs. Teoric ACF
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  acf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample ACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36),ylim=c(-1,1), 
       type="h",xlab="Lag",  ylab="", main="ACF Teoric")
  abline(h=0)
  
  #Sample PACF vs. Teoric PACF
  pacf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample PACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36, pacf=T),ylim=c(-1,1),
       type="h", xlab="Lag", ylab="", main="PACF Teoric")
  abline(h=0)
  par(mfrow=c(1,1))
}
################# Fi Validation #################################
```


### MA(1)


```{r}
(mod1=arima(d1d12lnserie,order=c(0,0,1),seasonal=list(order=c(1,0,1),period=12)))
mod1$coef/sqrt(diag(mod1$var.coef))
```

The intercept and sar1 not significant



```{r}
(mod1=arima(lnserie,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12)))
mod1$coef/sqrt(diag(mod1$var.coef))
```


### ARMA(1,1)

```{r}
(mod2=arima(d1d12lnserie,order=c(1,0,1),seasonal=list(order=c(1,0,1),period=12)))
coef(mod2)/sqrt(diag(mod2$var.coef))
```
Neither SAR1 or the intercept are significant. 


```{r}
(mod2=arima(lnserie,order=c(1,1,1),seasonal=list(order=c(0,1,1),period=12)))
coef(mod2)/sqrt(diag(mod2$var.coef))
```




### MA(2)

```{r}
(mod3=arima(d1d12lnserie,order=c(0,0,2),seasonal=list(order=c(0,0,1),period=12)))
coef(mod3)/sqrt(diag(mod3$var.coef))
```
We remove SAR1 and the intercept. 

```{r}
(mod3=arima(lnserie,order=c(0,1,2),seasonal=list(order=c(0,1,1),period=12)))
coef(mod3)/sqrt(diag(mod3$var.coef))
```




### ARMA(1,2)

```{r}
(mod4=arima(d1d12lnserie,order=c(1,0,2),seasonal=list(order=c(0,0,1),period=12)))

coef(mod4)/sqrt(diag(mod4$var.coef))
```
We remove SAR1 and the intercept. 

```{r}
(mod4=arima(lnserie,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12)))

coef(mod4)/sqrt(diag(mod4$var.coef))
```

*Note: Seasonal part = MA(1)

```{r}
library(kableExtra)
library(dplyr)
comp_models <- data.frame("Models" = c("MA(1)", "ARMA(1,1)", "MA(2)", "ARMA(1,2)"),
                          "Log-Likelihood" = c(mod1$loglik, mod2$loglik, mod3$loglik, mod4$loglik),
                          "AIC" = c(mod1$aic, mod2$aic, mod3$aic, mod4$aic))



best_rows <- order(comp_models$AIC)[1:2]

kbl(comp_models, booktabs = TRUE, caption = "Model Comparison") %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE) %>%
  row_spec(best_rows, bold = TRUE)
```

Based on the values of **log likelihood** y **AIC**, **MA(2)** , **ARMA(1,2)** are the two best performing models. 

## Validation

### MA(2)

```{r}
dades=d1d12lnserie
model=mod3
validation(model,dades)
```

*What to interpretate:*

*- Residuals: normality, homogeneity and independence*
*- Volatility (models for the variance)*
*- outlier detection*
*- Independence: Ljung box () and durbin watson test*
*- Discussing causality and invertibility based on polynomial values of MA*
*Compare sample ACF/PACF with theoretical ACF/PACF.*

### ARMA(1,2)

```{r}
dades=d1d12lnserie
model=mod4
validation(model,dades)
```

*What to interpretate:*

*- Residuals: normality, homogeneity and independence*
*- Volatility (models for the variance)*
*- outlier detection*
*- Independence: Ljung box () and durbin watson test*
*- Discussing causality and invertibility based on polynomial values of MA*
*Compare sample ACF/PACF with theoretical ACF/PACF.*

## Prediction assesment & Forecast

### MA(2)

```{r}
ultim=c(2018,12)
pdq=c(0,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```


```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2014,2020),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2014:2020,lty=3,col=4)
```

Model MA(2) is stable

```{r}
obs=window(serie,start=ultim)

mod.RMSE1=sqrt(sum((obs-pr)^2)/12)
mod.MAE1=sum(abs(obs-pr))/12
mod.RMSPE1=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE1,"MAE"=mod.MAE1,"RMSPE"=mod.RMSPE1,"MAPE"=mod.MAPE1)
```

```{r}
mCI1=mean(tu-tl)

cat("\nMean Length CI: ",mCI1)
```


```{r}
##### Long-term predictions with the complete model ######

pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

#Intervals
tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2015,2021), ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2015:2021,lty=3,col=4)
```

```{r}
(previs1=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```


### ARMA(1,2)


```{r}
ultim=c(2018,12)
pdq=c(1,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```

```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2014,2020),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2014:2020,lty=3,col=4)
```

Model ARMA(1,2) is stable

```{r}
obs=window(serie,start=ultim)

mod.RMSE1=sqrt(sum((obs-pr)^2)/12)
mod.MAE1=sum(abs(obs-pr))/12
mod.RMSPE1=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE1,"MAE"=mod.MAE1,"RMSPE"=mod.RMSPE1,"MAPE"=mod.MAPE1)
```

```{r}
mCI1=mean(tu-tl)

cat("\nMean Length CI: ",mCI1)
```

*comment on mean length of confidence interval of both models and how both are almost the same and also compare RMSE, MAE, RMSPE AND MAPE of both models*

```{r}
(previs=window(cbind(tl,pr,tu,serie,error=round(serie-pr,3)),start=ultim))

##### Long-term predictions with the complete model ######


pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

#Intervals
tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2015,2021),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2015:2021,lty=3,col=4)
```


```{r}
(previs2=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```

*Critically interpret the results and discuss model limitations in the context of the data.*
*Provide the explicit statistical expressions of the selected ARIMA models*

## ARIMAX


### Calendar effect


```{r}
source("CalendarEffects.r")
source("atipics2.R")
```



```{r}
data=c(start(lnserie)[1],start(lnserie)[2], length(lnserie))

(wTradDays=Wtrad(data)) 
(wEast=Weaster(data)) 



```



#### Cleaning on ARMA(1,2)



```{r}
(mod4ECWD=arima(lnserie,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),
                xreg=data.frame(wTradDays,wEast)))


```

```{r}
validation(mod4ECWD, d1d12lnserie)
```


```{r}
modEC.atip=outdetec(mod4ECWD,dif=c(1,12),crit=2.6,LS=T)

##Tabla de at?picos encontrados con su fecha
atipics=modEC.atip$atip[order(modEC.atip$atip[,1]),]
meses=c("Ene","Feb","Mar","Abr","May","Jun","Jul","Ago","Sep","Oct","Nov","Dic")

modEC.atip$sigma2

data.frame(atipics,Fecha=paste(meses[(atipics[,1]-1)%%12+1],start(serie)[1]+((atipics[,1]-1)%/%12)),perc.Obs=atipics[,3])
```

```{r}
lnserie.lin=lineal(lnserie,modEC.atip$atip) # linearized series: free of outliers

#Determine if variance(serie.lin) is constant and how many differences of seasonal and regular order need! In this case, variance constant, d=1, and D=1 with S=12

d1d12lnserie.lin=diff(diff(lnserie.lin,12))
```


```{r}
(modEC.lin=arima(lnserie.lin,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),xreg=data.frame(wTradDays,wEast)))
```

```{r}
validation(modEC.lin, diff(diff(lnserie.lin,12)))
```


```{r}
EfecTD=coef(modEC.lin)["wTradDays"]*wTradDays
EfecIN=coef(modEC.lin)["wEast"]*wEast
lnserieEC.lin=lnserie.lin-EfecTD-EfecIN
plot(lnserie-lnserieEC.lin)
```

```{r}
plot(lnserieEC.lin,col=2)
lines(lnserie)
```

```{r}
par(mfrow = c(1,2))
acf(d1d12lnserie.lin, main = "ACF", ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d1d12lnserie.lin, main = "PACF", ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
```


**Possible models**

- Regular
  - MA(1)
  - MA(3)
  - MA(11)
- Seasonal:
  - MA(1)


### Three possible models

**MA(1)**

```{r}
# (mod1x=arima(d1d12lnserie.lin,order=c(0,0,1),
#              seasonal=list(order=c(0,0,1),period=12), 
#              xreg = data.frame(wTradDays,wEast)))
# coef(mod1x)/sqrt(diag(mod1x$var.coef))
(mod1x=arima(lnserie.lin,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12),
             ,xreg=data.frame(wTradDays,wEast)))
coef(mod1x)/sqrt(diag(mod1x$var.coef))
```


```{r}
validation(mod1x, d1d12lnserie.lin)
```


**MA(2)**

```{r}
# (mod2x=arima(d1d12lnserie.lin,order=c(0,0,3),seasonal=list(order=c(0,0,1),period=12)))
# coef(mod2x)/sqrt(diag(mod2x$var.coef))
(mod2x=arima(lnserie.lin,order=c(0,1,2),seasonal=list(order=c(0,1,1),period=12),
             # fixed = c(NA,0,NA,NA),
             xreg=data.frame(wTradDays,wEast)))
coef(mod2x)/sqrt(diag(mod2x$var.coef))
```


```{r}
validation(mod2x, d1d12lnserie.lin)
```


**MA(11)**

```{r}
# (mod3x=arima(d1d12lnserie.lin,order=c(0,0,11),seasonal=list(order=c(0,0,1),period=12)))
# coef(mod3x)/sqrt(diag(mod3x$var.coef))
(mod3x=arima(lnserie.lin,order=c(0,1,11),seasonal=list(order=c(0,1,1),period=12),
             # fixed = c(NA,0,NA,0,0,NA,0,NA,0,0,NA,NA),
             xreg=data.frame(wTradDays,wEast)))
coef(mod3x)/sqrt(diag(mod3x$var.coef))
```


```{r}
validation(mod3x, d1d12lnserie.lin)
```


```{r}
(mod4x=arima(lnserie.lin,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),
             ,xreg=data.frame(wTradDays,wEast)))
coef(mod4x)/sqrt(diag(mod4x$var.coef))
```

```{r}
validation(mod4x, d1d12lnserie.lin)
```


**MA(7) **

```{r}
(mod5x=arima(lnserie.lin,order=c(1,1,5),seasonal=list(order=c(0,1,1),period=12),
             ,xreg=data.frame(wTradDays,wEast)))
coef(mod5x)/sqrt(diag(mod5x$var.coef))
```


```{r}
validation(mod5x, d1d12lnserie.lin)
```


**Next steps**

1. model MA(1), MA(3) and MA(11): evaluate AIC
2. Validation
3. Prediction assessment and forecasting

### Comparison of ARIMA models vs ARIMAX models

- AIC
- Prediction metrics (MAPE, RMSE, etc.)



