---
title: "Modelling gasoline consumption in Spain (1993-2019)"
subtitle: "Time Series MESIO UPC-UB"
date: today
author: "Abdelrahman Eissa & Benjamín Adasme"
format: 
  revealjs:
    echo: false
    embed-resources: true
execute: 
  warning: false
---

## Purpose and motivation: Why gasoline?

- Importance in global economy 
- Proxy for the national economy's health
- Diesel boom in Spain in the 90s
- Energy transition and decarbonization

## Objective {.center}

Capture the temporal structure of gasoline consumption, isolate exogenous effects (calendar/anomalies) and generate a robust model.

## Data in its historical context

```{r}
library(knitr)
library(kableExtra)
library(dplyr)
```


```{r}
serie=window(ts(read.table("Gasolina.dat"),start=1993,freq=12))
```

```{r}
plot(serie, main="Gasoline consumption in Spain", ylab="Thousands of tons")
abline(v=1993:2020,lty=3,col=4)
abline(h= 600)
```

## Methodology


- The Box-Jenkins iterative method: Identification, Estimation, Validation, Prediction Assessment and Forecasting
- From non-stationary to stationary series: logarithm, regular and seasonal differentiation.
- Calendar effects: importance for gasoline consumption estimation
- Outlier treatment: AO, TS, LS

## Results {.center}

## Model identification

```{r}
lnserie <- log(serie)
d1d12lnserie <- diff(diff(lnserie,12))

par(mfrow = c(1,2))
acf(d1d12lnserie, main = "ACF", ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d1d12lnserie, main = "PACF", ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
```

## Proposed ARIMA models

```{r}
mod1=arima(lnserie,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12))
mod2=arima(lnserie,order=c(1,1,1),seasonal=list(order=c(0,1,1),period=12))
mod3=arima(lnserie,order=c(0,1,2),seasonal=list(order=c(0,1,1),period=12))
mod4=arima(lnserie,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12))

comp_models <- tibble::tibble("Models" = c("MA(1)", "ARMA(1,1)", "MA(2)", "ARMA(1,2)"),
                          "Log-Likelihood" = round(c(mod1$loglik, mod2$loglik, mod3$loglik, mod4$loglik),2),
                          "AIC" = round(c(mod1$aic, mod2$aic, mod3$aic, mod4$aic),2),
                          "AR(1)" = round(c(NA, mod2$coef[1], NA, mod4$coef[1]),2),
                          "MA(1)" = round(c(mod1$coef[1], mod2$coef[2], mod3$coef[1], mod4$coef[2]),2),
                          "MA(2)" = round(c(NA,NA,mod3$coef[2], mod4$coef[3]),2),
                          "SMA(1)" = round(c(mod1$coef[2], mod2$coef[3], mod3$coef[3], mod4$coef[4]),2))

best_rows <- order(comp_models$AIC)[1:2]

kbl(comp_models, booktabs = TRUE, caption = "ARIMA Models Comparison") %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE) %>%
  row_spec(best_rows, bold = TRUE)
```

Seasonal MA(1). All coefficients are significant. Two models selected: MA(2) and ARMA(1,2)

## Prediction

```{r}
ultim=c(2018,12)
pdq=c(0,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)

par(mar=c(2,2,2,2),
    mfrow = c(2,1))
ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2014,2020),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2014:2020,lty=3,col=4)


## EVAL MOD3

obs=window(serie,start=ultim)

mod.RMSE1=sqrt(sum((obs-pr)^2)/12)
mod.MAE1=sum(abs(obs-pr))/12
mod.RMSPE1=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1=sum(abs(obs-pr)/obs)/12
mCI1=mean(tu-tl)

### Plot mod4

pdq=c(1,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)

ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2014,2020),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2014:2020,lty=3,col=4)

obs=window(serie,start=ultim)

mod.RMSE2=sqrt(sum((obs-pr)^2)/12)
mod.MAE2=sum(abs(obs-pr))/12
mod.RMSPE2=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE2=sum(abs(obs-pr)/obs)/12

mCI2=mean(tu-tl)
```


## 

ARIMA(1,1,2) performs better than ARIMA(0,1,2) on the classic ARIMA approach

```{r}
comp_pred <- tibble::tibble("Model" = c("ARIMA(0,1,2)", "ARIMA(1,1,2)"),
               "RMSE" = round(c(mod.RMSE1, mod.RMSE2),2),
               "MAE" = round(c(mod.MAE1, mod.MAE2),2),
               "RMSPE" = round(c(mod.RMSPE1, mod.RMSPE2),2),
               "MAPE" = round(c(mod.MAPE1, mod.MAPE2),2),
               "Mean CI" = round(c(mCI1, mCI2),2))

kable(comp_pred)

```

## Calendar effects and outlier detection {.small}

Outlier detection was performed on the SARIMA(1,1,2)(0,1,1) model to linearize the series. Also, it was included the calendar effect of working days and Easter holiday. 

```{r}
source("CalendarEffects.r")
source("atipics2.R")


data=c(start(lnserie)[1],start(lnserie)[2], length(lnserie))

wTradDays=Wtrad(data)
wEast=Weaster(data) 

mod4ECWD=arima(lnserie,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),
                xreg=data.frame(wTradDays,wEast))

modEC.atip=outdetec(mod4ECWD,dif=c(1,12),crit=2.6,LS=T)

##Tabla de at?picos encontrados con su fecha
atipics=modEC.atip$atip[order(modEC.atip$atip[,1]),]
# meses=c("Ene","Feb","Mar","Abr","May","Jun","Jul","Ago","Sep","Oct","Nov","Dic")

# out_dates <- data.frame(atipics,Fecha=paste(meses[(atipics[,1]-1)%%12+1],start(serie)[1]+((atipics[,1]-1)%/%12)),perc.Obs=atipics[,3])
# 
# out_dates$W_coeff <- round(out_dates$W_coeff,3)
# out_dates$ABS_L_Ratio <- round(out_dates$ABS_L_Ratio,3)
# out_dates$perc.Obs <- round(out_dates$perc.Obs,3)

serie.lin <- lineal(serie, modEC.atip$atip)
lnserie.lin=lineal(lnserie,modEC.atip$atip) # linearized series: free of outliers

#Determine if variance(serie.lin) is constant and how many differences of seasonal and regular order need! In this case, variance constant, d=1, and D=1 with S=12

d1d12lnserie.lin=diff(diff(lnserie.lin,12))

modEC.lin=arima(lnserie.lin,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),xreg=data.frame(wTradDays,wEast))

EfecTD=coef(modEC.lin)["wTradDays"]*wTradDays
EfecIN=coef(modEC.lin)["wEast"]*wEast
lnserieEC.lin=lnserie.lin-EfecTD-EfecIN

par(mar=c(rep(3,4)))
plot(lnserieEC.lin,col=2, main = "Outlier detection")
lines(lnserie)
legend("topright",
       legend = c("Original Series",
                  "Series without outliers"),
       col = c(1,2),
       lty = c(1,1))
```

## ARIMAX models

```{r}
par(mfrow = c(1,2))
acf(d1d12lnserie.lin, main = "ACF", ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d1d12lnserie.lin, main = "PACF", ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
```

Proposed models for clean series: Regular MA(1), MA(3), MA(7) and ARIMA(1,2) with Seasonal MA(1). 

## 

```{r}
mod1x=arima(lnserie.lin,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12),
             ,xreg=data.frame(wTradDays,wEast))
mod2x=arima(lnserie.lin,order=c(0,1,2),seasonal=list(order=c(0,1,1),period=12),
             # fixed = c(NA,0,NA,NA),
             xreg=data.frame(wTradDays,wEast))
mod3x=arima(lnserie.lin,order=c(0,1,11),seasonal=list(order=c(0,1,1),period=12),
             # fixed = c(NA,0,NA,0,0,NA,0,NA,0,0,NA,NA),
             xreg=data.frame(wTradDays,wEast))
mod4x=arima(lnserie.lin,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),
             ,xreg=data.frame(wTradDays,wEast))
mod5x=arima(lnserie.lin,order=c(1,1,5),seasonal=list(order=c(0,1,1),period=12),
             ,xreg=data.frame(wTradDays,wEast))


arimax_comp1 <- tibble::tibble(
  "ARIMAX Model" = c("MA(1)", "MA(2)", "MA(11)", "ARMA(1,2)", "ARMA(1,5)"),
  "Log-Likelihood" = round(c(mod1x$loglik, mod2x$loglik, mod3x$loglik, mod4x$loglik, mod5x$loglik),2),
  "AIC" = round(c(mod1x$aic, mod2x$aic, mod3x$aic, mod4x$aic, mod5x$aic),2)
)

best_rows <- order(arimax_comp1$AIC)[1:2]


kbl(arimax_comp1, booktabs = TRUE, caption = "ARIMAX Models Comparison") %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE) %>%
  row_spec(best_rows, bold = TRUE)
```

Selected ARIMAX models for prediction: MA(11) and ARMA(1,2)

##

```{r}
## MA 11
ultim=c(2018,12)
pdq=c(0,1,11)
PDQ=c(0,1,1)

# serie2=window(serie.lin,end=ultim)
lnserie2=window(lnserie.lin,end=ultim)
# serie1=window(serie.lin,end=ultim+c(1,0))
# lnserie1=window(lnserie.lin,end=ultim+c(1,0))

# modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12),
           xreg=data.frame(wTradDays[1:312],wEast[1:312]))

pred=predict(modB,n.ahead=12, 
             newxreg = data.frame(wTradDays[313:324],wEast[313:324]))
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)

obs=window(serie.lin,start=ultim)

mod.RMSE1x=sqrt(sum((obs-pr)^2)/12)
mod.MAE1x=sum(abs(obs-pr))/12
mod.RMSPE1x=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1x=sum(abs(obs-pr)/obs)/12
mCI1x=mean(tu-tl)

```

```{r}
## ARMA(1,2)
ultim=c(2018,12)
pdq=c(1,1,2)
PDQ=c(0,1,1)

# serie2=window(serie.lin,end=ultim)
lnserie2=window(lnserie.lin,end=ultim)
# serie1=window(serie.lin,end=ultim+c(1,0))
# lnserie1=window(lnserie.lin,end=ultim+c(1,0))

# modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12),
           xreg=data.frame(wTradDays[1:312],wEast[1:312]))

pred=predict(modB,n.ahead=12,
             newxreg = data.frame(wTradDays[313:324],wEast[313:324]))
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)

obs=window(serie.lin,start=ultim)

mod.RMSE2x=sqrt(sum((obs-pr)^2)/12)
mod.MAE2x=sum(abs(obs-pr))/12
mod.RMSPE2x=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE2x=sum(abs(obs-pr)/obs)/12
mCI2x=mean(tu-tl)
```





```{r}
arimax_comp2 <- tibble::tibble(
  "Model" = c("ARIMAX(0,1,11)", "ARIMAX(1,1,2)"),
  "RMSE" = round(c(mod.RMSE1x, mod.RMSE2x), 2),
  "MAE" = round(c(mod.MAE1x, mod.MAE2x), 2),
  "RMSPE" = round(c(mod.RMSPE1x, mod.RMSPE2x), 2),
  "MAPE" = round(c(mod.MAPE1x, mod.MAPE2x), 2),
  "Mean CI" = round(c(mCI1x, mCI2x), 2)
)

kbl(arimax_comp2, booktabs = TRUE, caption = "ARIMAX Models Prediction assessment") %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE)
```

## Final comparison

```{r}
precomp <- bind_rows(comp_pred, arimax_comp2)

final_comp <- tibble("Models" = c(comp_pred$Model, arimax_comp2$Model),
                     "Log-Likelihood" = c(comp_models$`Log-Likelihood`[3:4],
                                          arimax_comp1$`Log-Likelihood`[3:4]),
                     "AIC" = c(comp_models$AIC[3:4],
                                          arimax_comp1$AIC[3:4]))

final_comp <- final_comp |> bind_cols(precomp[,-1])

kbl(final_comp, booktabs = TRUE) |> 
  kable_styling(latex_options = "hold_position", full_width = FALSE) |> 
  row_spec(4, bold = T)

```


##


```{r}
#| fig-cap: "Final model forecast for the next 12 months"
#| fig-height: 5
ultim=c(2018,12)
pdq=c(1,1,2)
PDQ=c(0,1,1)

mod_final <- arima(lnserie.lin, order = pdq, 
                   seasonal = list(order=PDQ, seasonal=12),
                   xreg = data.frame(wTradDays, wEast))

newyear <- window(ts(rep(1,12),start=2020,freq=12))

data=c(start(newyear)[1],start(newyear)[2], length(newyear))

newWtrad <- Wtrad(data)
newEast <- Weaster(data)

pred=predict(mod_final,n.ahead=12,
             newxreg = data.frame(newWtrad, newEast))
pr<-ts(c(tail(lnserie.lin,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

#Intervals
tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

par(mar=rep(3,4))
ts.plot(serie.lin,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2015,2021), ylim=c(300,600),type="o",main=paste("Model ARIMAX(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12", "+CE",sep=""))
abline(v=2015:2021,lty=3,col=4)
```



## Conclusions {.smaller}

- **Initial Paradox Explained**: Although uncorrected ARIMA models can occasionally yield lower point-prediction errors due to sheer luck or overfitting to historical noise, they remain statistically invalid.
- **The ARIMAX Advantage**: Incorporating calendar effects and outliers via a SARIMAX(1,1,2) model properly addresses non-stationarity, successfully passes the Ljung-Box test, and eliminates residual patterns.
- **Uncertainty Halved**: By accounting for external variations simultaneously through exogenous regressors ($xreg$), the ARIMAX model slashes the prediction interval width nearly in half, drastically reducing forecasting uncertainty.
- **Strategic Future Projection**: For robust future forecasting, this framework allows the model to seamlessly adapt to upcoming calendar structures while safely neutralizing past anomalous shocks.


