---
title: "Report Gasoline"
subtitle: "Time Series - MESIO UPC-UB"
author: "Abdelrahman Eissa & Benjamín Adasme"
date: "2026-05-18"
output: 
  pdf_document: 
    toc: true
    toc_depth: 4
    df_print: kable
    fig_height: 4
---

\newpage

```{r include=FALSE}
knitr::opts_chunk$set(echo = F, 
                      message = F,
                      warning = F)
```

## Introduction

- Gasoline consumption in Spain and the world: factors
- Dataset Gasoline
- Objectives 

## Methodology

Explain the methodology (steps, approaches, etc. )

- Classic ARIMA: Box-jenkins method for time series: transform into stationary, (P)ACF, propose models and evaluate them.
  -Identification
  - Estimation
  - Diagnosis and validation
  - Prediction and forecasting
- ARIMAX: calendar effects and outlier treatment: clean by calendar effect and outlier detection -> repeat process of classic ARIMA

## Results and discussion

### Classic ARIMA approach

### Model identification

```{r}
serie=window(ts(read.table("Gasolina.dat"),start=1993,freq=12))
```

```{r}
plot(serie, main="Gasoline consumption in Spain", ylab="Thousands of tons")
abline(v=1993:2020,lty=3,col=4)
abline(h= 600)
```

The model has a mean of `r round(mean(serie),2)` and a variance of `r round(sd(serie),2)`. Since we detect this time serie is not stationary, we applied the next transformations:

- Box-Cox transformation: 
- Seasonality:
- Non-constant variance: 


```{r, fig.height=3, fig.cap="Non-constant variance serie"}
par(mfrow = c(1,2),
    mar=c(4,3,3,2))
boxplot(serie~floor(time(serie)), main="a) Boxplot by year", xlab="Year")
m=apply(matrix(serie,nr=12),2,mean)
v=apply(matrix(serie,nr=12),2,var)
plot(m,v,xlab="Medias anuales",ylab="Varianzas anuales",main="b) Anual mean and variance")
abline(lm(v~m),col=2,lty=3,lwd=2)

```

```{r, fig.height=4.5, fig.width=6, fig.cap="Decomposition of log-transformed series"}
lnserie <- log(serie)
plot(decompose(lnserie))
```
```{r}
d1d12lnserie <- diff(diff(lnserie,12))
plot(d1d12lnserie, main=paste0("Stationary series - Mean: ", round(mean(d1d12lnserie),2)))
abline(h = mean(d1d12lnserie), col = 2, lty = 2)
```



### Estimation for plausible models

```{r}
par(mfrow = c(1,2))
acf(d1d12lnserie, main = "ACF", ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d1d12lnserie, main = "PACF", ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
```

Seasonal part: MA (1)

Regular part: 

- MA(1)
- ARMA(1,1)
- MA(2)
- ARMA(1,2)

```{r}
mod1=arima(lnserie,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12))
mod2=arima(lnserie,order=c(1,1,1),seasonal=list(order=c(0,1,1),period=12))
mod3=arima(lnserie,order=c(0,1,2),seasonal=list(order=c(0,1,1),period=12))
mod4=arima(lnserie,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12))

comp_models <- tibble::tibble("Models" = c("MA(1)", "ARMA(1,1)", "MA(2)", "ARMA(1,2)"),
                          "Log-Likelihood" = round(c(mod1$loglik, mod2$loglik, mod3$loglik, mod4$loglik),2),
                          "AIC" = round(c(mod1$aic, mod2$aic, mod3$aic, mod4$aic),2),
                          "AR(1)" = round(c(NA, mod2$coef[1], NA, mod4$coef[1]),2),
                          "MA(1)" = round(c(mod1$coef[1], mod2$coef[2], mod3$coef[1], mod4$coef[2]),2),
                          "MA(2)" = round(c(NA,NA,mod3$coef[2], mod4$coef[3]),2),
                          "SMA(1)" = round(c(mod1$coef[2], mod2$coef[3], mod3$coef[3], mod4$coef[4]),2))

comp_models
```

### Model validation

```{r}

mini_val <- function(model, datos){
  par(mfrow = c(2,2), mar=c(3,3,3,3))
  resid <- resid(model)
  #Residuals plot
  plot(resid,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resid),3*sd(resid)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resid)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resid)
  qqline(resid,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resid,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resid),sd=sd(resid)),col=2,add=T)
}


mini_val(mod3, d1d12lnserie)

```


```{r}
mini_val(mod4, d1d12lnserie)
```


### Prediction assesment and forecasting

- Both models are stable

**MA(2) stability**

```{r}
ultim=c(2018,12)
pdq=c(0,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

(ma_est <- tibble::tibble("Variables" = names(modA$coef),
                         "Full" = round(coef(modA),2),
                         "Cut" = round(coef(modB),2),
                         "Diference" = abs(Full - Cut)))

```




**ARMA(1,2) stability** 

```{r}
ultim=c(2018,12)
pdq=c(1,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

(arma_est <- tibble::tibble("Variables" = names(modA$coef),
                         "Full" = round(coef(modA),2),
                         "Cut" = round(coef(modB),2),
                         "Diference" = abs(Full - Cut)))



```

```{r, fig.height=5}

### MA(2) plot

pdq=c(0,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)

par(mar=c(2,2,2,2),
    mfrow = c(2,1))
ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2014,2020),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2014:2020,lty=3,col=4)


## EVAL MOD3

obs=window(serie,start=ultim)

mod.RMSE1=sqrt(sum((obs-pr)^2)/12)
mod.MAE1=sum(abs(obs-pr))/12
mod.RMSPE1=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1=sum(abs(obs-pr)/obs)/12
mCI1=mean(tu-tl)

### Plot mod4

pdq=c(1,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)

ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2014,2020),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2014:2020,lty=3,col=4)

obs=window(serie,start=ultim)

mod.RMSE2=sqrt(sum((obs-pr)^2)/12)
mod.MAE2=sum(abs(obs-pr))/12
mod.RMSPE2=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE2=sum(abs(obs-pr)/obs)/12

mCI2=mean(tu-tl)
```


```{r}
comp_pred <- tibble::tibble("Model" = c("MA(2)", "ARMA(1,2)"),
               "RMSE" = round(c(mod.RMSE1, mod.RMSE2),2),
               "MAE" = round(c(mod.MAE1, mod.MAE2),2),
               "RMSPE" = round(c(mod.RMSPE1, mod.RMSPE2),2),
               "MAPE" = round(c(mod.MAPE1, mod.MAPE2),2),
               "Mean CI" = round(c(mCI1, mCI2),2))

comp_pred
```

**Selected model: ARMA(1,2)**

```{r}
## Prediction modelo final:

pdq=c(1,1,2)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12))
modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12))

pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


# (previs=window(cbind(tl,pr,tu,serie,error=round(serie-pr,3)),start=ultim))

##### Long-term predictions with the complete model ######


pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

#Intervals
tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

par(mar=c(rep(3,4)))
ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(2015,2021),ylim=c(300,600),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=2015:2021,lty=3,col=4)
```



### ARIMAX Analysis: Outlier and Calendar Effects Treatment

```{r}
source("CalendarEffects.r")
source("atipics2.R")
library(broom)
```

```{r}
data=c(start(lnserie)[1],start(lnserie)[2], length(lnserie))

wTradDays=Wtrad(data)
wEast=Weaster(data) 

(mod4ECWD=arima(lnserie,order=c(1,1,2),seasonal=list(order=c(0,1,1),period=12),
                xreg=data.frame(wTradDays,wEast)))

mod4ECWD$coef/sqrt(diag(mod4ECWD$var.coef))

tidy(mod4ECWD) 
```

```{r}
modEC.atip=outdetec(mod4ECWD,dif=c(1,12),crit=2.6,LS=T)

##Tabla de at?picos encontrados con su fecha
atipics=modEC.atip$atip[order(modEC.atip$atip[,1]),]
meses=c("Ene","Feb","Mar","Abr","May","Jun","Jul","Ago","Sep","Oct","Nov","Dic")

modEC.atip$sigma2

out_dates <- data.frame(atipics,Fecha=paste(meses[(atipics[,1]-1)%%12+1],start(serie)[1]+((atipics[,1]-1)%/%12)),perc.Obs=atipics[,3])

out_dates$W_coeff <- round(out_dates$W_coeff,3)
out_dates$ABS_L_Ratio <- round(out_dates$ABS_L_Ratio,3)
out_dates$perc.Obs <- round(out_dates$perc.Obs,3)

out_dates
```


### ARIMA and ARIMAX models comparison

## Conclusions

\newpage

## Appendices

### Validation function

```{r validation function 2, echo=TRUE}
#################Validation#################################
validation=function(model,dades){
  s=frequency(get(model$series))
  resid=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resid,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resid),3*sd(resid)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resid)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resid)
  qqline(resid,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resid,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resid),sd=sd(resid)),col=2,add=T)
  
  
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resid,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #ACF & PACF of square residuals 
  par(mfrow=c(1,2))
  acf(resid^2,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid^2,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=36)
  names(psis)=paste("psi",1:36)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:20])
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=36)
  names(pis)=paste("pi",1:36)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:20])
  
  ## Add here complementary tests (use with caution!)
  ##---------------------------------------------------------
  cat("\nNormality Tests\n")
  cat("\n--------------------\n")
 
  ##Shapiro-Wilks Normality test
  print(shapiro.test(resid(model)))

  suppressMessages(require(nortest,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  print(ad.test(resid(model)))
  
  suppressMessages(require(tseries,quietly=TRUE,warn.conflicts=FALSE))
  ##Jarque-Bera test
  print(jarque.bera.test(resid(model)))
  
  cat("\nHomoscedasticity Test\n")
  cat("\n--------------------\n")
  suppressMessages(require(lmtest,quietly=TRUE,warn.conflicts=FALSE))
  ##Breusch-Pagan test
  obs=get(model$series)
  print(bptest(resid(model)~I(obs-resid(model))))
  
  cat("\nIndependence Tests\n")
  cat("\n--------------------\n")
  
  ##Durbin-Watson test
  print(dwtest(resid(model)~I(1:length(resid(model)))))
  
  ##Ljung-Box test
  cat("\nLjung-Box test\n")
  print(t(apply(matrix(c(1:4,(1:4)*s)),1,function(el) {
    te=Box.test(resid(model),type="Ljung-Box",lag=el)
    c(lag=(te$parameter),statistic=te$statistic[[1]],p.value=te$p.value)})))
  

  #Sample ACF vs. Teoric ACF
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  acf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample ACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36),ylim=c(-1,1), 
       type="h",xlab="Lag",  ylab="", main="ACF Teoric")
  abline(h=0)
  
  #Sample PACF vs. Teoric PACF
  pacf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample PACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36, pacf=T),ylim=c(-1,1),
       type="h", xlab="Lag", ylab="", main="PACF Teoric")
  abline(h=0)
  par(mfrow=c(1,1))
}
################# Fi Validation #################################
```

### ARIMA models validation detailed

#### MA(2)

```{r}
validation(mod3, d1d12lnserie)
```

#### ARMA(1,2)

```{r}
validation(mod4, d1d12lnserie)
```

